\exercise{Chain and cycle\label{exMasterChain}}{2}
Let’s study another biased random walk.
\subquestion
Consider a walker on a 3-chain, (1)-(2)-(3). When in a node $i<3$, the walker moves to node $i+1$ at rate $2$. When in a node $i>1$ it moves to node $i-1$ at rate $1$. Determine the probabilities of finding the walker in node $i$ in the stationary state of the system.

\solution 
The system has the Jacobian
\eq{
{\bf J}= \aveccc{ -2 & 1 & 0 \\ 2 & -3 & 1 \\ 0 & 2 & -1}
}
Because the system is a tree, we can use the detailed balance condition. The flow on the link connecting node 1 and 2 are 
\eqa{
j_{21}&=&j_{1\to 2}-j_{2\to 1} \\
  &=& 2v_1 - v_2  
}
Setting $j_{21}=0$ reveals 
\eq{
v_2 = 2v_1
}
Consider the link from 2 to 3 gives us 
\eqa{
j_{32} &=& j_{2\to 3} - j_{3\to 2} \\
  &=& 2v_2 -v_3
}
and hence 
\eq{
v_3=2v_2 = 4 v_1
}
Thus we have the vector 
\eq{
\vec{v} = \avec{v_1 \\ 2v_1 \\ 4v_1}
}
The sum of the elements is $7v_1$ and hence we have to set $v_1=1/7$ for the elements to add up to 1. To be sure we can confirm
\eq{
\aveccc{ -2 & 1 & 0 \\ 2 & -3 & 1 \\ 0 & 2 & -1}\avec{1/7\\ 2/7\\ 4/7} = \avec{0 \\ 0 \\ 0}
}
so the steady state proabilities are 
\eq{
p_1 = \frac{1}{7},\quad p_2=\frac{2}{7}, \quad p_3=\frac{4}{7}.
}

\subquestion 
Now close the chain to form a triangle. In addition to the transitions from part (a), a walker that is in node 3 moves directly to node 1 at rate 2 and a walker that is in node 1 moves directly to node 3 at rate 1. Again find the stationary state.

\solution
Now the Jacobian is 
\eq{
{\bf J}= \aveccc{ -3 & 1 & 2 \\ 2 & -3 & 1 \\ 1 & 2 & -3}
}
The easiest way to the solution is to spot the kernel vector by visual inspection. Since the column sums are identical $(1,1,1)^{\rm T}$ must be an eigenvector and multiplying 
\eq{
\aveccc{ -3 & 1 & 2 \\ 2 & -3 & 1 \\ 1 & 2 & -3}\avec{1\\ 1\\ 1} = \avec{0\\ 0 \\ 0}
}
reveals that it is the desired kernel vector. Noramlizing it leads us directly to the result 
\eq{
p_1 = \frac{1}{3},\quad p_2=\frac{1}{3}, \quad p_3=\frac{1}{3}.
}
If you can't spot the eigenvector directly, you can always go back to the nodewise eqations by multiplying the Jacobian with an arbitrary vector $\vec{v}$ to find the conditions
\eq{
\aveccc{ -3 & 1 & 2 \\ 2 & -3 & 1 \\ 1 & 2 & -3}\avec{v_1\\ v_2\\ v_3} = \avec{-3v_1+v_2+2v_3\\ -3v_2 + 2v_1 + v_3  \\ -3v_3 + v_1 + 2v_2}
}
We can now normalize $v_1=1$ and consider the first condition 
\eqa{
0 &=& -3v_1+v_2+2v_3 \\  
  &=& -3+v_2+2v_3
}
which yields 
\eq{
v_2=3-2v_3
}
We now consider the second condition 
\eqa{
0 &=& -3v_2 + 2v_1 + v_3 \\
  &=& -3(3-2v_3) + 2 +v_3 \\
  &=& -9+6v_3+2+v_3 \\
  &=& 7v_3-7
}
which gives us 
\eq{
v_3 = 1 
}
and hence 
\eq{
v_2 = 1
}
so we arrive at
\eq{
\vec{v} = (1\\ 1\\ 1)^{\rm T}
}
this way, before normalizing to find the result 
\eq{
p_1 = \frac{1}{3},\quad p_2=\frac{1}{3}, \quad p_3=\frac{1}{3}.
}
as above. 

\subquestion
Show that the detailed balance condition is met in stationary state in the network from (a) but not in the network from (b).

\solution 
We have used the detailed balance assumption in (a) to find the solution and then verified that the solution is correct, which proed that the system from (a) meets the detailed balance condition. 

To check the detailed balance condition in (b) we compute the net flux between node 1 and 2 
\eqa{
J_{21} &=& j_{1\to 2} - j_{2\to 1} \\
 &=& 2v_1 - v_2 = 2/3-1/3 = 1/3
}
There is a residual flow of 1/3 walker per unit of time through the link, which stems from a circular flow of the walker around the cycle. This violates the detailed balance assumption.