\exercise{Matrix perturbation theory\label{exCentralImport}}{4}
Let's suppose we make a small change to our Jacobian such that ${\bf J}={\bf A}+p{\bf B}$, where ${\bf A}$ is the original Jacobian, ${\bf B}$ describes  the change and $p$ is a parameter that is initially zero. Let's assume we know the eigenvalues $\lambda_n$ and the left eigenvectors $\vec{w_n^\dag}$ and the right eigenvectors $\vec{v_n}$ of $J$ for $p=0$. We now compute how the eigenvalues change as we increase $p$:

\subquestion 
Start with the (right) eigenvector equation for $\bf J$, substitute $\bf J$ and differentiate the whole equation with respect to $p$. Finally eliminate $\bf A$ by using the substitution backwards.  
\solution 
We start with 
\eq{
{\bf J} \vec{v_n} = \lambda_n \vec{v_n}
}
Substituting 
\eq{
{\bf J}={\bf A}+p{\bf B} 
}
yields 
\eq{
({\bf A}+p{\bf B}) \vec{v_n} = \lambda_n \vec{v_n}
}
When we change $p$ both the eigenvectors and eigenvalues will change so we need a to use a product rule in the differentiation 
\eqa{
\partial_p ({\bf A}+p{\bf B}) \vec{v_n} &=& \partial_p (\lambda_n \vec{v_n}) \\ 
{\bf B}\vec{v_n} + {\bf J}\vec{v_n}'   &=&  \lambda_n' \vec{v_n} + \lambda_n \vec{v_n}'
}
In the second row of this equation I used a dash to indicate the derivative with respect to $p$. I have also already put the $\bf J$ back in by undoing the substitution in the second row.

\subquestion You should now have one equation that contains the change of $\partial_p \lambda_n$ but also $\partial_p \vec{v_n}$, which we don't want to compute. To fix this multiply your equation with $\vec{w_n^\dag}$ from the left, then find a way to eliminate $\partial_p \vec{v_n}$, and solve for $\partial_p \lambda_n$.  

\solution 
Multiplying $\vec{w_n^\dag}$ from the left yields
\eq{
\vec{w_n^\dag} {\bf B}\vec{v_n} + \vec{w_n^\dag} {\bf J}\vec{v_n}'   =   \vec{w_n^\dag} \lambda_n' \vec{v_n} + \vec{w_n^\dag} \lambda_n \vec{v_n}'
}
Our goal is to remove the $\vec{v_n}'$ which on both side of the equation. This makes you wonder if they maybe cancel each other, doesn't it? Recall that
\eq{
    \vec{w_n^\dag} {\bf J} = \lambda_n \vec{w_n^\dag}.
}
Substituting this relationship yields 
\eq{
\vec{w_n^\dag} {\bf B}\vec{v_n} + \vec{w_n^\dag} \lambda_n \vec{v_n}'   =   \vec{w_n^\dag} \lambda_n' \vec{v_n} + \vec{w_n^\dag} \lambda_n \vec{v_n}'
}
Now the terms are indeed identical and we can cancel them which leads to  
\eq{
\vec{w_n^\dag} {\bf B}\vec{v_n}    =   \vec{w_n^\dag} \lambda_n' \vec{v_n}
}
Since $\lambda_n'$ is a number we can just pull it out from between the two vectors and solve for 
\eq{
\frac{\partial \lambda_n}{\partial p} = \lambda_n' = \frac{\vec{w_n^\dag}{\bf B} \vec{v_n}}{\vec{w_n^\dag}\vec{v_n}}.
}
\subquestion Does the result give you an idea for another way in which we could rank the importance of nodes?

\solution 
The equation gives us a way to estimate the effect of the eigenvalues if we change the matrix. Suppose I wanted to improve the stability of the system and I had a way of adding a small controller to one node, which provides a little bit of negative feedback without changing the steady state. In that case I might want to rank the nodes in terms ot their effect on the eigenvalue. 

If I chose to control node $i$, the matrix $\bf B$ has only one non-zero element  $B_{ii}=-1$, and hence we can compute the impact on the rightmost eigenvalue as 
\eq{
\frac{\partial \lambda_1}{\partial p} = - \frac{w_{1,i}^\dag v_{1,i}}{\vec{w_1^\dag}\vec{v_1}}.
}
which acts as a kind of `rightmost eigenvector perturbation centrality'.
