\exercise{The Zombie Apocalypse ... will be weird}{3}
During the zombie apocalypse each zombie apocalypse each zombie is killed within a day, but not before infecting 3 people with the zombie virus. Within a day the infected people turn to zombies. 

\subquestion Model the system as a discrete-time map where the $Z_t$ and $I_t$ are the numbers of zombies and infected on day $t$.

\solution 
Using the information from the question, we can model the system by the discrete time map
\eqa{
Z_{t+1} &=& I_{t} \\
I_{t+1} &=& 3 Z_{t} 
}

\subquestion Use the method from the chapter to compute $Z_t$,  $I_t$ for all $t>0$.

\solution
We define the vector 
\eq{\vec{X_t}=\avec{Z_t\\ I_t}.}
We can now write the system in matrix
\eq{
\vec{X_{t+1}} = {\bf J} \vec{X_t,}
}
where 
\eq{
{\bf J}= \avecc{0 & 1 \\ 3 & 0}.
}
To compute the eigenvectors of $\bf J$ we solve 
\eq{
\lambda \avec{1\\x} = \avecc{0 & 1 \\ 3 & 0} \avec{1\\x}
}
which we can also write as
\eqa{
\lambda & = &  x \\ 
\lambda x & = & 3
}
Substituting the first line into the second one gives us 
\eq{
x^2 = 3
}
and hence
\eq{
x=\pm \sqrt{3}.
}
Using $\lambda=x$ we have now two eigenvalues
\eq{
\lambda_1 = \sqrt{3}, \qqq \lambda_2 = -\sqrt{3}
}
and the corresponding eigenvectors 
\eq{
\vec{v_1} = \avec{1 \\ \sqrt{3}},\qqq \vec{v_1} = \avec{1 \\ -\sqrt{3}}.     
}
Therefore the general solution is 
\eq{
\avec{Z_t \\ I_t} = c_1 (\sqrt{3})^t \avec{1 \\ sqrt{3}} + c_2 (-\sqrt{3})^t \avec{1 \\ sqrt{3}} \avec{1 \\ -sqrt{3}}
}

\subquestion Using your solution, consider the case where the outbreak starts with one infected on day $t=0$ and compute the number of zombies on days 1,2,20, and 21.

\solution
To find the particular solution we need to determine $c_1$ and $c_2$ by solving the system on day 0,
\eq{
\avec{0 \\ 1} = c_1  \avec{1 \\ sqrt{3}} + c_2 \avec{1 \\ \sqrt{-3}}.
}
The upper line reads 
\eq{
0= c_1 + c_2 
}
hence 
\eq{
c_1=-c_2.
}
Substituting this into the lower line yields
\eq{
1 = c_1 \sqrt{3} + c_1 \sqrt{3} = 2c_1 \sqrt{3} 
}
hence 
\eq{
c_1 = \frac{1}{2\sqrt{3}} \qqq c_2 = - \frac{1}{2\sqrt{3}}
}
We now have the particular solution 
\eq{
\avec{Z_t \\ I_t} = \frac{1}{2\sqrt{3}} (\sqrt{3})^t \avec{1 \\ sqrt{3}} - \frac{1}{2\sqrt{3}} (-\sqrt{3})^t \avec{1 \\ sqrt{3}}
}
We can write the line for the Zombies as
\eq{
Z_t= \frac{1}{2\sqrt{3}} (\sqrt{3})^t - \frac{1}{2\sqrt{3}} (-\sqrt{3})^t
}
by pulling a factor $(-1)^t$ from the second term we can further simplify this to 
\eq{
Z_t= \frac{1}{2\sqrt{3}} \left( (\sqrt{3})^t - (-1)^t  (\sqrt{3})^t\right) 
}
and by cancelling a factor of $\sqrt{3}$ and collecting the terms we arrive at 
\eq{
Z_t= \frac{1-(-1)^t}{2} \sqrt{3})^{t-1}. 
}
Form this form it is now apparent that the number of zombies will be 0 on all even days. On odd days first factor becomes 1 and we have 
\eq{
Z_t= \sqrt{3})^{t-1} \qqq \mbox{[On odd days]} 
}
Hence the number of zombies is zero on day 2 and day 20. On day 1 it is $\sqrt{3}^0=1$ and on day 21 it is $\sqrt{3}^20 =3^10$ which is 59049. Of course it will be zero again on day 22. 
