\exercise{Two faces of heterogeneity\label{exCowsabstractGC}}{2}
Estimate the size of the giant component in the following networks:

\subquestion An ER-graph with $z=2$ with 100,000 nodes.

\solution
We know that the giant component size in an ER network obeys
\eq{
s = 1-\exp{-sz}
}
For $z=2$ we solve this by iteration, which yields $s\approx 0.79681$ as the fraction of nodes in the giant component. To get the number we multiply by $N$ which yields $79,681$ nodes. 

[There is an 11\% chance that a randomly picked node is not in the giant component. All other components are small in comparison, containing only a few nodes. Hence the expected number of nodes that we find in the component of a randomly picked nodes is  
approximately $71,000$ but this wasn't the question.] 

\subquestion A network consisting of 50,000 nodes of degree 1 and 50,000 nodes of degree 3.

\solution
The degree distribution is 
\eq{p_k= \frac12 (\delta_{k,1} + \delta_{k,3})}

\paragraph{Mean degree.}We compute the mean degree
\eq{
z=\sum k p_k = \frac{1}{2}(1+3) = 2
}

\paragraph{Excess degree distribution}This leads to the excess degree distribution
\eq{
q_k = \frac{(k+1)p_{k+1}}{z} = \frac{1}{4} (\delta_{k,0} + 3 \delta_{k,2})
}
\paragraph{Digression: Mean excess degree.}Let us also compute the mean excess degree, we don't actually need it here but it will be interesting to compare it to the mean excess degree in part (c).
\eq{
q=\sum k q_k = \sum k \frac{1}{4} (\delta_{k,0} + 3 \delta_{k,2}) = 
 \frac{0 + 6}{4} = 1.5
}
\paragraph{Giant component size.}To compute the probability that a link does not lead to the giant component, we use the self-consistency equation
\eqa{
v&=&\sum q_k v^k  \\
 &=&\sum \frac{1}{4} (\delta_{k,0} + 3 \delta_{k,2}) v^k \\
 &=& \frac{1 + 3 v^2}{4}
}
We can write this as quadratic polynomial
\eq{
0 = 3 v^2 - 4v +1.  
}
We could now solve this in the usual way (e.g.~completing the square). A more insightful solution is to realize that we know one solution already as $v=1$ is always a solution due to the nature of the self-consistency approach. We are not interested in this solution itself but we can divide the corresponding factor $(v-1)$ out of the polynomial, using polynomial long division:
\eq{
(3v^2-4v+1)/(v-1)=3v-1 
}
and hence the solution that we are actually interested in is 
\eq{
v=\frac{1}{3}
}
[There will actually a longer exercise on the polynomial long dividion in Chap.~10]

Now that we know $v$ we use the equation for the giant component size 
\eqa{
s&=&1-\sum p_k v^k \\
 &=&1-\sum \frac12 (\delta_{k,1} + \delta_{k,3}) v^k \\
 &=&1- \frac{v+v^3}{2} \\
 &=&1- \frac{1/3+1/27}{2} \\
 &=& \frac{54}{54} - \frac{9}{54} - \frac{1}{54}\\
 &=&  \frac{22}{27} = 0.\xoverline{814} 
}
In other words we expect 81.481 nodes to be in the giant component. 

\subquestion A network of 50,000 nodes with degree 1 and 50,000 that follow a Poisson distribution with mean 3.

\solution
We can write the degree distribution as 
\eq{
p_k=\frac{1}{2}\delta_{k,1}+\frac{1}{2}\frac{\exp{-3}3^k}{k!}.
}
\paragraph{Mean degree.}We can compute the mean degree 
\eqa{
 z&=&\sum k p_k \\
  &=&\sum k \left(\frac{1}{2}\delta_{k,1}+\frac{1}{2}\frac{\exp{-3}3^k}{k!} \right) \\
  &=& \frac{1}{2} + \frac{1}{2}\exp{-3}  \sum k \frac{3^k}{k!} \\
  &=& \frac{1}{2} + \frac{1}{2}\exp{-3} \sum_{k=1}^\infty \frac{3^k}{(k-1)!} \\
  &=& \frac{1}{2} + \frac{1}{2}\exp{-3} \sum_{k=0}^\infty \frac{3^(k+1)}{k!} \\
  &=& \frac{1}{2} + \frac{3}{2}\exp{-3} \sum \frac{3^k}{k!} \\
  &=& \frac{1}{2} + \frac{3}{2}\exp{-3} \exp{3} \\
  &=& \frac{1}{2} + \frac{3}{2} = 2 \\
}
\paragraph{Excess degree distribution.}OK, we could have just guessed this result, but this was a good opportunity to practice some of the tools in the toolkit in a situation where the result was easy to guess (and hence errors easy to spot). 
In any case, the next step in our work flow is to construct the excess degree distribution
\eqa{
q_k&=&\frac{(k+1)p_{k+1}}{z} \\
  &=& \frac{1}{2} \left( \frac{k+1}{2}\delta_{k+1,1}+\frac{k+1}{2}\frac{\exp{-3}3^{k+1}}{(k+1)!} \right) \\
  &=& \frac{1}{2} \left( \frac{1}{2}\delta_{k+1,1}+\frac{1}{2}\frac{\exp{-3}3^{k+1}}{k!} \right) \\
  &=&  \frac{1}{4}\delta_{k,0}+\frac{3}{4}\frac{\exp{-3}3^{k}}{k!}  
}
\paragraph{Digression: Mean excess degree in heterogeneous networks.}We can now compute the mean excess degree to compare it to the mean excess degree from part (b)
\eqa{
q&=&\sum k q_k \\ 
 &=&\sum k \left( \frac{1}{4}\delta_{k,0}+\frac{3}{4}\frac{\exp{-3}3^{k}}{k!} \right) \\
 &=&\frac{1}{4} \left( k\delta_{k,0} \right) +\frac{3}{4}\left( k\frac{\exp{-3}3^{k}}{k!} \right) \\
 &=& \frac{3}{4}\left( \sum k\frac{\exp{-3}3^{k}}{k!} \right) \\
 &=& \frac{3}{4}\exp{-3}\left( k\frac{3^{k}}{k!} \right) \\
 &=& \frac{9}{4} = 2.25 \\
}
In the last step we remembered 
\eq{
\sum k\frac{3^{k}}{k!} = 3 \exp{3} 
}
which we saw in our calculation of the mean degree. (if you don't remember this, just cancel the factor $k$ in the sum with the last $k$ from the factorial, shift the index, pull a factor of 3 out of the sum, use the exponential series definition and you will arrive at the same result.)  

The mean excess degree of this networks turns out to be quite a bit higher than the mean excess degree of $q=1.5$ that we found for the network in Ex.~8.3b. This actually makes sense because the network here is more heterogeneous, and we know that heterogeneity drives the mean excess degree. Now the interesting question is, does this also mean that the heterogeneous network has a larger giant component? High mean excess degree is good for giant components isn't it? 

\paragraph{Giant component size.}To find the giant component size we compute $v$ from 
\eqa{
v&=&\sum q_k v^k \\
 &=& \sum \left(\frac{1}{4}\delta_{k,0}+\frac{3}{4}\frac{\exp{-3}3^{k}}{k!} \right)v^k \\  
  &=& \left(\frac{1}{4}\sum \delta_{k,0}v^k\right)+\left(\frac{3}{4}\exp{-3}\sum\frac{3^{k}}{k!}v^k \right) \\
    &=& \left(\frac{1}{4}\sum \delta_{k,0}v^0\right)+\left(\frac{3}{4}\exp{-3}\sum\frac{(3v)^{k}}{k!} \right) \\
    &=& \frac{1}{4}+\frac{3}{4}\exp{-3}\exp{3v}  \\
    &=& \frac{1}{4}+\frac{3}{4}\exp{3(v-1)}   
}
We solve this equation by iteration. Because we found $v=1/3$ in a very similar network in Ex.~8.3b, we use this as the starting value. This yields the sequence 
\begin{center}
    1/3, 0.35150146242, 0.35718728095, 0.35903130571, 0.35963614646, 0.35983526429, 0.35990089436, 0.3599225349, 0.35992967149, 0.35993202508.      
\end{center}
At this point I am convinced that the result will be very close to $v=0.36$. So lets use this in the final step. We compute the giant component size as 
\eqa{
s&=&1-\sum p_k q^k \\
 &=&1-\sum \left( \frac{1}{2}\delta_{k,1}+\frac{1}{2}\frac{\exp{-3}3^k}{k!}\right) v^k \\ 
 &=&1-\frac{1}{2}\left(\sum \delta_{k,1}v^k\right)+\frac{1}{2}\exp{-3}\left(\sum \frac{3^k}{k!}\right) v^k \\ 
 &=&1-\frac{1}{2}v+\frac{1}{2}\exp{-3}\left(\sum \frac{(3v)^k}{k!}\right) \\ 
 &=&1-\frac{1}{2}v+\frac{1}{2}\exp{-3}\exp{3v} \\
 &=&1-\frac{1}{2}v+\frac{1}{2}\exp{3(v-1)} \\
}
In principle we could now just substitute $v$ and be done with it, but a more elegant way is to consider our final equation for $v$ before the iteration
\eq{
v=\frac{1}{4}+\frac{3}{4}\exp{3(v-1)} 
}
we can solve for the exponential 
\eq{
\exp{3(v-1)}=\frac{4v-1}{3} 
}
substituting back into our equation for $s$ we have
\eq{
s=1-\frac{1}{2}v+\frac{1}{2}\frac{4v-1}{3}=1-\frac{7v-1}{6} 
}
If you wonder whether we got this right you can quickly check that $v=1$ would result in $s=0$, which it still does. To find the solution we substitute $v=0.36$ which gives us 
\eq{
s=0.74\xoverline{6}
}
so roughly 75\%. This is a bit smaller than the 84\% that we found 
in the similar network in Ex.~8.3b. In comparison the network in this exercise is a bit more heterogeneous. 
\paragraph{Conclusion.}This is interesting because it shows heterogeneity is double-edged sword. on the one hand heterogeneous networks have a higher mean excess degree which makes it easier to form a giant component. However, heterogeneity also means that we have a more nodes of low degree and therefore smaller giant component sizes.   

