\exercise{Attacks on networks}{4}
In the next chapter we study attacks on networks. This exercise offers a little preview of some relevant calculations. Let us consider a network where every node has degree 4 before an attack removes half the links at random.  

\subquestion Construct $p_k$, the probability that a node has $k$ links after the attack. Then use the degree distribution $p_k$ to construct the degree generating function $G_{\rm a}$ and the excess degree generating function $Q_{\rm a}$ after the attack.

\solution
\paragraph{Degree distribution.}We have to work out the probabilities for all of the different scenarios: The probability that a node survives with all links intact is $1/16$, the probability that a node loses exactly one link is $4/16$ (see Chap.~7), and so on. This leads to the following degree distribution
\eq{
p_k = \frac{1}{16} (\delta_{k,0}+4\delta_{k,1}+6\delta_{k,2}+4\delta_{k,3}+\delta_{k,4})   
}
\paragraph{Degree genereating function after the attack.}Now, we can calculate the degree generating function $G_{\rm a}$ after the attack using the definition of the generating function for a degree distribution
\eq{
G=\sum p_k x^k.
}
In this case the generating fuction after the attack is
\eqa{
G_{\rm a}&=&\frac{1}{16}(1+4x+6x^2+4x^3+x^4)\\
&=& \left(\frac{1}{2}+\frac{x}{2}\right)^4 
}
\paragraph{Mean degree.}When we remove half the links from the network we expect the degree to halve as well (remember $z=2K/N$? If we reduce $K$, $z$ shrinks proportionally) 
To verify we can compute

\eqa{
z&=&\sum k p_k \\
&=& \sum \frac{1}{16} (k\delta_{k,0}+4k\delta_{k,1}+6k\delta_{k,2}+4k\delta_{k,3}+k\delta_{k,4}) \\
&=& \frac{1}{16} (4+6\cdot 2+4\cdot 3+4) = \frac{32}{16}=2.  
}
\paragraph{Excess degree distribution.}We can now compute the excess degree distribution in the usual way 
\eqa{
q_k&=&\frac{(k+1)p_{k+1}}{z} \\
   &=& \frac{1}{2}\frac{1}{16} (4\delta_{k+1,1}+12\delta_{k+1,2}+12\delta_{k+1,3}+4\delta_{k+1,4}) \\
   &=& \frac{1}{8} (\delta_{k,0}+3\delta_{k,1}+3\delta_{k,2}+1\delta_{k,3}) \\
}
\paragraph{Excess degree generating function after the attack.}Finally, we compute the excess degree generating function $Q_{\rm a}$, using the defintion
\eq{
Q=\sum q_k x^k.
}
We find the solution
\eqa{
Q_{\rm a}&=&\frac{1}{8}(1+3x+3x^2+x^3)\\
&=&\left(\frac{1}{2}+\frac{x}{2}\right)^3.
}
This wasn't horrible, but it certainly wasn't an elegant calculation either. 

\subquestion Now we do the same calculation with generating functions: Let $G$ be the degree distribution before the attack. Also write the generating function $A$ for a coin flip that gives us a result of 0 or 1 with equal probability. Can you compute $G_{\rm a}$ and $Q_{\rm a}$ from $G$ and $A$?

\solution
This is sooooo much nicer. The generating function before the attack is 
\eq{
G=x^4.
}
The generating function for a coin flip that results in 0 or 1 is
\eq{
A=\frac{1}{2}+\frac{x}{2}.
}
\paragraph{Generating function after the attack.}To find the generating function after the attack we can think along the following lines. If we pick a node at random then the number of links that the node had in the original network is described by $G$. However, since some of the links may have been destroyed in the attack we need to flip a coin for each link to see if it still there and then sum over the results of the coin flips. This scenario is covered by our dice-of-dice rule so the result is 
\eq{
G_{\rm a}=G(A)=\left(\frac{1}{2}+\frac{x}{2}\right)^4 
}
\paragraph{Excess degree distribution after the attack.}We can now compute the excess degree distribution after the attack using the convenient formula from this chapter
\eq{
Q_{\rm a}=\frac{G'_{\rm a}}{G'_{\rm a}(1)} 
}
We compute 
\eq{
G'_{\rm a}=  4\left(\frac{1}{2}+\frac{x}{2}\right)^3 \frac{1}{2} = 2 \left(\frac{1}{2}+\frac{x}{2}\right)^3
}
which reveals the mean degree after the attack $G'_{\rm a}(1)=2$ and furthermore 
\eq{
Q_{\rm a}= \left(\frac{1}{2}+\frac{x}{2}\right)^3.
}

This calculation seems almost magical: At first it is hard to understand why we do these specific steps, but once we have gained this understanding we can do great things with little effort. 

\solutionend
