\exercise{Population growth \label{exGrowPopulation}}{2}
\label{exGrowthQuadratic}
Sexual reproduction requires that two individuals of a species meet. If we apply the law of mass action naively we may arrive at a model of the form
$\dot{x}=x^2$. Solve the initial value problem for $x(0)=1$ and explore what happens. 

\solution
Starting from 
\eq{
\ddt x = x^2
}
separation of variables leads to
\eq{
   \frac{1}{x^2} \d{x} = \dt.
}
Integrating we find 
\eqa{
\int \frac{1}{x^2} \d{x} & = & \int \dt \\
  - \frac{1}{x} &=& t + C
}
where $C$ is the constant of integration. Solving for $x$ yields 
\eq{
   x(t) = - \frac{1}{t+C}.
}
This is already the general solution to the problem. To find the particular solution that solves the initial value problem we use 
\eq{
  x(0) = 1 = - \frac{1}{C},
}
which we can solve for 
\eq{
    C=-1.
}
Hence the solution to the initial value problem is 
\eq{
    x(t) = - \frac{1}{t-1} = \frac{1}{t-1}.
}
From this solution we can see that something interesting is happening here. At $t=1$ the denominator of the fraction is zero and the population size $x$ diverges toward $+\infty$. This is indeed a correct result, quadratic growth leads to an explosion that reaches $\infty$ in finite time. 

We can see from this result that geometric growth ($\dot{x}\sim x^\gamma$ with $\gamma>1$) is much much more violent than exponential growth, which also reaches $\infty$, but only after infinite time. 

For biological populations quadratic growth is a bad model for reasons that we discover in Chap.~17.  